---
title: "NBA_Draft_Model_2018"
author: "Ryan Salsbury"
date: "5/13/2018"
output: html_document
---

Test various models to predict the NBA performance of the upcoming 2018 draft class and the probability of them becoming a superstar, star, starter, bench bust player at the next level

Load Packages
```{r}

library(sqldf)
library(caTools)
library(Hmisc)
library(rpart)
library(caret)
library(randomForest)

```

Linear Regression Model - Predict OWS
```{r}
Training <- read.csv("Training.csv")
Test <- read.csv("Test.csv")
Prospects <- read.csv("Prospects.csv")

#Get all numerical input variables and output variable(NBA_OWS)
Training <- Training[, c(2:50,52)]
Test <- Test[, c(2:50,52)]
Prospects <- Prospects[, c(2:50)]

#Scale Data
Training[c(1:49)] = scale(Training[c(1:49)])
Test[c(1:49)] = scale(Test[c(1:49)])

#Create model using all variables
lrOWSmodel1 = lm(NBA_OWS ~., data = Training)

#view model summary and understand important variables
summary(lrOWSmodel1)

#Create new model using most important variables
lrOWSmodel2 = lm(NBA_OWS ~ NCAA_AdjTempo + NCAA_MIN_PG + NCAA_STL_PG + NCAA_FT_FGA + NCAA_TSpct + NCAA_PPS + Height + NCAASeasons + AgeAtDraft + NCAA_WS, data = Training)

#view new model summary 
summary(lrOWSmodel2)

#View accuracy of model on predicting test data using RMSE
RMSE(predict(lrOWSmodel2, Test), Test$NBA_OWS)

#RMSE = 3.15






```

Linear Regression Model - Predict DWS
```{r}
Training <- read.csv("Training.csv")
Test <- read.csv("Test.csv")
Prospects <- read.csv("Prospects.csv")

#Get all numerical input variables and output variable(NBA_DWS)
Training <- Training[, c(2:50,53)]
Test <- Test[, c(2:50,53)]
Prospects <- Prospects[, c(2:50)]

#Scale Data
Training[c(1:49)] = scale(Training[c(1:49)])
Test[c(1:49)] = scale(Test[c(1:49)])

#Create model using all variables
lrDWSmodel1 = lm(NBA_DWS ~., data = Training)

#view model summary and understand important variables
summary(lrDWSmodel1)

#Create new model using most important variables
lrDWSmodel2 = lm(NBA_DWS ~ + NCAA_WS + NCAA_STL_PG + NCAA_DblDbl + NCAA_TplDbl + NCAA_PPS + Height + NCAA_DRtg + AgeAtDraft + NCAASeasons, data = Training)

#view new model summary 
summary(lrDWSmodel2)
#Adj. R-Squared=.38

#View accuracy of model on predicting test data using RMSE
RMSE(predict(lrDWSmodel2, Test), Test$NBA_DWS)

#RMSE = 1.93

```


Decision Tree Model - Predict OWS
```{r}

#Load Data
Training <- read.csv("Training.csv")
Test <- read.csv("Test.csv")
Prospects <- read.csv("Prospects.csv")

#Remove non-numerical columns
Training <- Training[, c(2:50,52)]
Test <- Test[, c(2:50,52)]
Prospects <- Prospects[, c(2:50)]

#Create model using all variables
dtOWSmodel1 = rpart(formula = NBA_OWS ~ .,
                  data = Training)

#View details of model
summary(dtOWSmodel1)

#See how well it predicts test data
RMSE(predict(dtOWSmodel1, Test), Test$NBA_OWS)

#RMSE=2.39

#Modify model using only significant data
dtOWSmodel2 = rpart(formula = NBA_OWS ~ NCAA_WS + AgeAtDraft + NCAA_STL_PG + NCAA_DWS + NCAASeasons + NCAA_DRtg + NCAA_W + NCAA_PER + NCAA_Stl_TO,
                  data = Training)

#View details of model
summary(dtOWSmodel2)

#See how well new model predicts test data
RMSE(predict(dtOWSmodel2, Test), Test$NBA_OWS)

#RMSE=2.49

#view diagram of decision tree
plot(dtOWSmodel2)
text(dtOWSmodel2)

```

Decision Tree Model - Predict DWS
```{r}

#Load Data
Training <- read.csv("Training.csv")
Test <- read.csv("Test.csv")
Prospects <- read.csv("Prospects.csv")

#Remove non-numerical columns
Training <- Training[, c(2:50,53)]
Test <- Test[, c(2:50,53)]
Prospects <- Prospects[, c(2:50)]

#Create model using all variables
dtDWSmodel1 = rpart(formula = NBA_DWS ~ .,
                  data = Training)

#View details of model
summary(dtDWSmodel1)

#See how well it predicts test data
RMSE(predict(dtDWSmodel1, Test), Test$NBA_DWS)

#RMSE=2.14

#Modify model using only significant data
dtDWSmodel2 = rpart(formula = NBA_DWS ~ NCAA_WS + AgeAtDraft + NCAA_BLK_PG + NCAA_DWS + NCAASeasons
                                     + NCAA_DRtg + NCAA_DblDbl + NCAA_PF_PG + NCAA_DEF_PG,
                  data = Training)

#View details of model
summary(dtDWSmodel2)

#See how well new model predicts test data
RMSE(predict(dtDWSmodel2, Test), Test$NBA_DWS)

#RMSE=1.72

#view diagram of decision tree
plot(dtDWSmodel2)
text(dtDWSmodel2)

```

Random Forest - Predict OWS
```{r}
#Load data
Training <- read.csv("Training.csv")
Test <- read.csv("Test.csv")
Prospects <- read.csv("Prospects.csv")

#Get all numerical input variables and output variable(NBA_OWS)
Training <- Training[, c(3:6, 9:50,52)]
Test <- Test[, c(3:6, 9:50,52)]
Prospects <- Prospects[, c(3:6, 9:50)]

#Create Random Forest using all variables
rfOWSmodel1 = randomForest(x = Training[, c(1:46)],
                          y = Training$NBA_OWS,
                          ntree = 500, mtry = 46)

#view list of all variables and how important they are in predicting OWS
importance(rfOWSmodel1)

#Grab the 10 most important variables and output variable
rfTrainOWS <- sqldf("select NCAA_WS, NCAA_AdjTempo, NCAA_AdjOE, AgeAtDraft, NCAA_FTpct_PG, NCAA_STL_PG, NCAA_PPS, NCAA_PPR, NCAA_PER, NCAA_Ortg, NBA_OWS FROM Training")

rfTestOWS <- sqldf("select NCAA_WS, NCAA_AdjTempo, NCAA_AdjOE, AgeAtDraft, NCAA_FTpct_PG, NCAA_STL_PG, NCAA_PPS, NCAA_PPR, NCAA_PER, NCAA_Ortg, NBA_OWS FROM Test")

#Run the model with updated variables using 500 trees
rfOWSmodel2 = randomForest(x = rfTrainOWS[, c(1:10)],
                          y = rfTrainOWS$NBA_OWS,
                          ntree = 500, mtry = 10)

#view model details and calculate accuracy(RMSE) on Test Data
rfOWSmodel2
RMSE(predict(rfOWSmodel2, rfTestOWS), rfTestOWS$NBA_OWS)

#RMSE = 2.13 - Most Accurate Model

#Make predictions on new data(2018 Prospects)
rfProspectsOWS <- sqldf("select NCAA_WS, NCAA_AdjTempo, NCAA_AdjOE, AgeAtDraft, NCAA_FTpct_PG, NCAA_STL_PG, NCAA_PPS, NCAA_PPR, NCAA_PER, NCAA_Ortg FROM Prospects")

rfOWSpred = predict(rfOWSmodel2, newdata = as.data.frame(rfProspectsOWS))

#Convert predictions to data frame
rfOWSpred = as.data.frame(rfOWSpred)

#Read Prospects to get player names
Prospects <- read.csv("Prospects.csv")

#Join predictions to Prospects
ProspectOWSPred <- sqldf("select Prospects.Player, rfOWSpred.* from Prospects JOIN rfOWSpred ON Prospects.rowid = rfOWSpred.rowid")

```

Random Forest - Predict DWS
```{r}

#Load data
Training <- read.csv("Training.csv")
Test <- read.csv("Test.csv")
Prospects <- read.csv("Prospects.csv")

#Get all numerical input variables and output variable(NBA_DWS)
Training <- Training[, c(3:6, 9:50,53)]
Test <- Test[, c(3:6, 9:50,53)]
Prospects <- Prospects[, c(3:6, 9:50)]

#Create Random Forest using all variables
rfDWSmodel1 = randomForest(x = Training[, c(1:46)],
                          y = Training$NBA_DWS,
                          ntree = 500, mtry = 46)

#view list of all variables and how important they are in predicting DWS
importance(rfDWSmodel1)

#Grab the most important variables and the output variable
rfTrainDWS <- sqldf("select NCAA_WS, NCAA_DWS, NCAA_AdjOE, AgeAtDraft, NCAA_STL_PG, NCAA_PER, NCAA_PF_PG, NBA_DWS FROM Training")

rfTestDWS <- sqldf("select NCAA_WS, NCAA_DWS, NCAA_AdjOE, AgeAtDraft, NCAA_STL_PG, NCAA_PER, NCAA_PF_PG, NBA_DWS FROM Test")

#Run the model with updated variables using 500 trees
rfDWSmodel2 = randomForest(x = rfTrainDWS[, c(1:7)],
                          y = rfTrainDWS$NBA_DWS,
                          ntree = 500, mtry = 7)

#view model details and calculate accuracy(RMSE) on Test Data
rfDWSmodel2
RMSE(predict(rfDWSmodel2, rfTestDWS), rfTestDWS$NBA_DWS)

#RMSE = 1.64 - Most accurate Model

#Make predictions on new data(2018 Prospects)
rfProspectsDWS <- sqldf("select NCAA_WS, NCAA_DWS, NCAA_AdjOE, AgeAtDraft, NCAA_STL_PG, NCAA_PER, NCAA_PF_PG FROM Prospects")

rfDWSpred = predict(rfDWSmodel2, newdata = as.data.frame(rfProspectsDWS))

#Convert predictions to data frame
rfDWSpred = as.data.frame(rfDWSpred)

#Read Prospects to get player names
Prospects <- read.csv("Prospects.csv")

#Join predictions to Prospects
ProspectDWSPred <- sqldf("select Prospects.Player, rfDWSpred.* from Prospects JOIN rfDWSpred ON Prospects.rowid = rfDWSpred.rowid")

#Join DWS Predictions to OWS Predictions to get combined OWS and DWS Predictions
ProspectPredictions <- sqldf("select ProspectOWSPred.Player, ProspectOWSPred.rfOWSpred OWS, ProspectDWSPred.rfDWSpred DWS, ProspectOWSPred.rfOWSpred + ProspectDWSPred.rfDWSpred WS from ProspectOWSPred JOIN ProspectDWSPred ON ProspectOWSPred.Player = ProspectDWSPred.Player")

```

Predict Star Potential - Random Forest
```{r}

#Load Classification datasets
TrainingClass <- read.csv("TrainingClass.csv")
TestClass <- read.csv("TestClass.csv")
Prospects <- read.csv("Prospects.csv")

#Reduce Dataset so that the large number of bust players does not negatively impact results.
TrainingClass <- sqldf("select * from TrainingClass where rowid <= 280")

#Reduce dataset to numerical variables and outcome variable(Role)
TrainingClass <- TrainingClass[, c(3:6, 9:50,56)]
TestClass <- TestClass[, c(3:6,9:50,56)]
Prospects <- Prospects[, c(2:50)]

rfTrainRole <- sqldf("select * FROM TrainingClass")
rfTestRole <- sqldf("select * FROM TestClass")

#Convert outcome variable to a factor
rfTrainRole$Role <- as.factor(rfTrainRole$Role)
rfTestRole$Role <- as.factor(rfTestRole$Role)

#Create model with all variables
rfRolemodel = randomForest(x = rfTrainRole[, c(1:46)],
                          y = rfTrainRole$Role,
                          ntree =500, mtry=46)

#View variable importance
importance(rfRolemodel)

#Grab most important variables and output variable
rfTrainRole <- sqldf("select AgeAtDraft, NCAA_DWS, NCAA_AdjOE, NCAA_STL_PG,NCAA_AdjTempo, NCAA_PPR, NCAA_PER, NCAA_WS,NCAA_TOV_PG, NCAA_FTM_PG, NCAA_OFF_PG, Role FROM TrainingClass")

rfTestRole <- sqldf("select AgeAtDraft, NCAA_DWS, NCAA_AdjOE, NCAA_STL_PG,NCAA_AdjTempo, NCAA_PPR, NCAA_PER, NCAA_WS,NCAA_TOV_PG, NCAA_FTM_PG, NCAA_OFF_PG,Role FROM TestClass")

#convert outcome variable to factor
rfTrainRole$Role <- as.factor(rfTrainRole$Role)
rfTestRole$Role <- as.factor(rfTestRole$Role)


#Create new model with new variables using 100 trees and 8 variables at each split
rfRolemodel = randomForest(x = rfTrainRole[, c(1:11)],
                          y = rfTrainRole$Role,
                          ntree =100, mtry=8)

#view model details 
rfRolemodel

#Make predictions on new data(2018 Prospects)
rfProspectRole <- sqldf("select AgeAtDraft, NCAA_DWS, NCAA_AdjOE, NCAA_STL_PG,NCAA_AdjTempo, NCAA_PPR, NCAA_PER, NCAA_WS,NCAA_TOV_PG, NCAA_FTM_PG, NCAA_OFF_PG FROM Prospects")

#type = "prob"" gives the probabilty of each potential outcome rather than making a single prediction.
rfRolepred = predict(rfRolemodel, newdata = as.data.frame(rfProspectRole), type = "prob")

#Convert to data frame
rfRolepred = as.data.frame(rfRolepred)

#Update Column Name
colnames(rfRolepred) = c("Superstar", "Star", "Starter", "Bench", "Bust")

#Read Prospects to get player names
Prospects <- read.csv("Prospects.csv")

#Join Role Classification predictions to WS Predictions
ProspectRolePred <- sqldf("select * from Prospects JOIN rfRolepred ON Prospects.rowid = rfRolepred.rowid")


#Join Role Classification predictions to WS Predictions
ProspectPred <- sqldf("select ProspectPredictions.*, ProspectRolePred.Superstar, ProspectRolePred.Star, ProspectRolePred.Starter, ProspectRolePred.Bench, ProspectRolePred.Bust  from ProspectPredictions JOIN ProspectRolePred ON ProspectPredictions.Player = ProspectRolePred.Player")


```
